Type Checker Generation using Reference Attribute Grammars

### 2023-11-02 MSc.tex
Add some code samples and descriptions

\documentclass[nofilelist]{cslthse-msc}
% to show a list of used packages at the end of the document, delete the nofilelist option
%\documentclass{cslthse-msc}
\usepackage[utf8]{inputenc}
\usepackage[main=british,swedish]{babel}
\usepackage{amsmath}
%\usepackage{amsfonts}
%%\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{makeidx}
\usepackage{graphicx}
\usepackage[titletoc, header, page]{appendix}
\usepackage{transparent}
\usepackage{tikz}
\usepackage{enumitem}
\usetikzlibrary{fit,positioning,shapes.geometric}

@@

\chapter{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
%\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach. Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
\section{Results} % contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews). Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
\subsection{Input/output examples, with descriptions} %TODO
Bool language
Type variables
\section{Research questions}
\subsection{Research q 1}
\subsection{Research q 2}
\subsection{Research q 3}
List of challenges % TODO
  Splitting code between multiple JastAdd components (and java) can get messy
  Using JastAdd simultaneously as a library and a build component

>>>>

\documentclass[nofilelist]{cslthse-msc}
% to show a list of used packages at the end of the document, delete the nofilelist option
%\documentclass{cslthse-msc}
\usepackage[utf8]{inputenc}
\usepackage[main=british,swedish]{babel}
\usepackage{amsmath}
%\usepackage{amsfonts}
%%\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage{makeidx}
\usepackage{graphicx}
\usepackage[titletoc, header, page]{appendix}
\usepackage{transparent}
\usepackage{tikz}
\usepackage{enumitem}
\usepackage{listings}
\usetikzlibrary{fit,positioning,shapes.geometric}

\lstdefinestyle{mystyle}{
    basicstyle=\ttfamily,
    breakatwhitespace=false,
    breaklines=true,
    captionpos=b,
    keepspaces=true,
    numbers=left,
    numbersep=5pt,
    showspaces=false,
    showstringspaces=false,
    showtabs=false,
    tabsize=2
}
\lstset{style=mystyle}

@@

\chapter{Evaluation} % is the part where you present the finds. Depending on the area this part contains a subset or all of the following:
%\section{Experimental Setup} % should describe the details of the method used to evaluate your solution(s)/approach. Sometimes this is already addressed in the \textbf{Method}, sometimes this part replaces \textbf{Method}.
\section{Results} % contains the data (as tables, graphs) obtained via experiments  (benchmarking, polls, interviews). Here you should also describe the individual tables or graphs in text, pointing out interesting outliers and trends.
%\subsection{Input/output examples, with descriptions}
\subsection{Bools}

\begin{figure}[h]
\begin{lstlisting}[]
Expression ::= Term;

abstract Term;
True  : Term;
False : Term;
Or    : Term ::= Left:Term Right:Term;
Zero  : Term;
\end{lstlisting}
  \caption{AST specification for the Bools language}
  \label{boolsast}
\end{figure}
\begin{figure}[h]
\begin{lstlisting}[]
[T-True]
-----------
True : Bool

[T-False]
------------
False : Bool

[T-Or]
left : Bool,
right : Bool
----------------------
Or(left, right) : Bool

[T-Zero]
----------
Zero : Int
\end{lstlisting}
  \caption{Typing rules for the Bools language}
  \label{boolstr}
\end{figure}

\begin{figure}[h]
\begin{lstlisting}[]
aspect TypingRules {

syn Type Zero.type() {
return new Int();}
syn Type Or.type() {
ASTNode left = getChild(0);
ASTNode right = getChild(1);

if(left.type().matches(new Bool()) && right.type().matches(new Bool())) {
return new Bool();
}
throw new RuntimeException("Typechecking failed");
}
syn Type True.type() {
return new Bool();}
syn Type False.type() {
return new Bool();}
}
\end{lstlisting}
  \caption{Generated typing rules for the Bools language}
  \label{boolstrgen}
\end{figure}
A simple example of a language is provided in figure \ref{boolsast}, consisting of three values, \verb|True|, \verb|False| and \verb|Zero|; and the boolean operator \verb|Or| with two parameters, Left and Right.
In the AST specification, there are no limitations on what terms can be provided as parameters to the \verb|Or| operator, allowing illogical constructs such as \verb|Or(True, Zero)|.

To prevent cases like these, we define typing rules via the specification in figure \ref{boolstr}.
The typing rules for \verb|True|, \verb|False| and \verb|Zero| contain no premises and define the type as \verb|Bool| and \verb|Int| as appropriate.
The typing rule for the \verb|Or| operator, which takes two parameters \verb|left| and \verb|right|, contains two premises, specifying that the \verb|Or| operator has the type \verb|Bool| only if both the \verb|left| and \verb|right| terms also have the type \verb|Bool|.

The typing rules from figure \ref{boolstr} generate the type definitions in \ref{boolstrgen}.
The simple rule definitions for the values compile into equally simple implementations, consisting of a single return statement of a newly constructed \verb|Type| object.
For the \verb|Or| node, the code becomes slightly more complex.
First two variables are defined, corresponding to the parameters in the typing rule, and assigned to the children with the index corresponding to the position among the parameters.
This is followed by an if-statement, recusively evaluating the children's types and comparing the result the type demanded in the premise.
If the condition is true, a \verb|Bool| type is returned.
If not, a runtime exception is thrown to convey a type error.

\subsection{If-else extension}

\begin{figure}[h]
\begin{lstlisting}[]
IfElse : Term ::= Cond:Term Then:Term Else:Term;
\end{lstlisting}
  \caption{Extension for the Bools AST}
  \label{ifelseast}
\end{figure}
\begin{figure}[h]
\begin{lstlisting}[]
[T-IfElse]
x: Bool,
y: a,
z: a
------------------
IfElse(x, y, z): a
\end{lstlisting}
  \caption{Extension for the Bools typing rules with an if-else term}
  \label{ifelsetr}
\end{figure}
\begin{figure}[h]
\begin{lstlisting}[]
syn Type IfElse.type() {
ASTNode x = getChild(0);
ASTNode y = getChild(1);
ASTNode z = getChild(2);

Type tyvar_a = null;
if(tyvar_a == null) tyvar_a = y.type();
else if (!tyvar_a.matches(y.type()))
throw new RuntimeException("Typechecking failed: Type variable mismatch");
if(tyvar_a == null) tyvar_a = z.type();
else if (!tyvar_a.matches(z.type()))
throw new RuntimeException("Typechecking failed: Type variable mismatch");
if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
return tyvar_a;
}
throw new RuntimeException("Typechecking failed");
}
\end{lstlisting}
  \caption{Generated typing rules for the if-else term}
  \label{boolstrgen}
\end{figure}

%\begin{figure}
%\centering
%  \begin{minipage}[b]{.4\textwidth}
%  \centering
%\begin{lstlisting}[]
%[T-IfElse]
%x: Bool,
%y: a,
%z: a
%---------
%IfElse(x, y, z): a
%\end{lstlisting}
%  \captionof{figure}{A figure}
%  \label{fig:test1}
%\end{minipage}%
%\begin{minipage}[b]{.6\textwidth}
%  \centering
%\begin{lstlisting}[]
%syn Type IfElse.type() {
%ASTNode x = getChild(0);
%ASTNode y = getChild(1);
%ASTNode z = getChild(2);
%
%Type tyvar_a = null;
%if(tyvar_a == null) tyvar_a = y.type();
%else if (!tyvar_a.matches(y.type()))
%throw new RuntimeException("Typechecking failed: Type variable mismatch");
%if(tyvar_a == null) tyvar_a = z.type();
%else if (!tyvar_a.matches(z.type()))
%throw new RuntimeException("Typechecking failed: Type variable mismatch");
%if(x.type().matches(new Bool()) && y.type().matches(tyvar_a) && z.type().matches(tyvar_a)) {
%return tyvar_a;
%}
%throw new RuntimeException("Typechecking failed");
%}
%\end{lstlisting}
%  \captionof{figure}{Another figure}
%  \label{fig:test2}
%\end{minipage}
%\end{figure}
Extend bool with if-statement to show type variables
\section{Research questions}
\subsection{Research q 1}
\subsection{Research q 2}
\subsection{Research q 3}
List of challenges % TODO
  Splitting code between multiple JastAdd components (and java) can get messy
  Using JastAdd simultaneously as a library and a build component
